An objective assessment of point scoring systems in Formula 1 motor racing

### 2022-11-28 inst/jebo_F1points.tex
issue #185

%% This file is to be used as a template for your submission.
%% Rename this file and replace the text with the text of
%% your manuscript.
%%
%% The standard LaTeX document class "article" is recommended.
%% Use options letterpaper and 12pt.
\documentclass[a4,12pt]{article}

%% This is the recommended preamble for your document.

%% Load De Gruyter specific settings

\usepackage{amsmath}

%% The mathptmx package is recommended for Times compatible math symbols.
%% Use mtpro2 or mathtime instead of mathptmx if you have the commercially
%% available MathTime fonts.
%% Other options are txfonts (free) or belleek (free) or TM-Math (commercial)
\usepackage{mathptmx}

%% Use the graphics package to include figures
\usepackage{graphics}

%% Use natbib with these recommended options
\usepackage[authoryear,comma,longnamesfirst,sectionbib]{natbib}

\usepackage{setspace}\doublespacing

@@

\begin{titlepage}
  \begin{center}{\Large
      An objective assessment of point scoring systems in Formula 1 motor racing}\\
    name redacted\\
{\tt email redacted }\\
affiliation redacted
\end{center}

\end{titlepage}

@@

\newpage
\section*{An objective assessment of point scoring systems in Formula 1 motor racing}
\section*{Abstract}
%The points scoring system of Formula 1 motor racing is a long-standing
%contention among fans and competitors, with changes being
%controversial.  The inaugural points system was $(8,6,4,3,2)$, that
%is, 8 points to the winner, 6 to second place, and so on and in 2020
%the points system was changed to $(25,18,15,12,10,8,6,4,2,1)$.  The
%overall ranking of the competitors is determined by summing points
%awarded over a season, but this system suffers from two defects:
%firstly, the points are intrinsically arbitrary in the sense that the
%points system does not affect the competitors' behaviour but can
%change the overall ranking.  Secondly, it is difficult to assess
%statistical nulls using accumulated points.  Here, I use Plackett-Luce
%likelihood to identify a ranking for the competitors, this method
%including no arbitrary points system and being amenable to statistical
%testing of nulls.  I go on to assess a number of reasonable points
%systems objectively, by comparing likelihood-based rankings with
%points-based rankings.  In one well-defined statistical sense a Borda
%points system is preferred, and in another a ``winner takes all''
%system is best.

The points scoring system of Formula 1 motor racing is a long-standing
contention among fans and competitors.  The inaugural points system
was $(8,6,4,3,2)$, that is, 8 points to the winner, 6 to second place,
and so on; in 2020 the points system was changed to
$(25,18,15,12,10,8,6,4,2,1)$.  However, it is difficult to assess
statistical nulls using accumulated points.  Here, I use Plackett-Luce
likelihood to identify a ranking for the competitors, this method
being amenable to statistical testing.  I go on to assess a
number of reasonable points systems objectively: in one well-defined
statistical sense a Borda points system is optimal.

\newpage

\section{Introduction}\label{introduction}
Formula 1 motor racing is an important and prestigious motor sport
\citep{codling2017,jenkins2010}.  Season ranking is based on a points
allocation system wherein competitors are awarded points based on race
finishing order; points accumulate additively.  The overall
competition winner is the competitor who accumulates the most points
after the final race.

@@

(observe how this likelihood function, while informative about
$p_4+p_5$, is uninformative about $p_4\left|p_4+p_5\right.$). We now
use a technique due to \cite{hankin2010,hankin2017} and introduce
fictional (reified) entities whose nonzero Bradley-Terry strength
helps certain competitors or sets of competitors under certain
conditions. The canonical example would be the home-ground advantage
in association football.  If players (teams) $1,2$ with strengths
$p_1,p_2$ compete, and if our observation were $a$ home wins and $b$
away wins for team $1$, and $c$ home wins and $d$ away wins for
team~$2$, then a suitable likelihood function would be

@@

\noindent where $p_H$ is a quantification of the beneficial home
ground effect.  Similar techniques have been used to account for the
first-move advantage in chess, and effective coordination between
members of doubles tennis teams; we may use a similar device to
account for (e.g.) wet conditions in Formula 1.  Here I analyse
seasons 2016-2019 using the \texttt{hyper2} package \citep{hankin2017}
which implements the Plackett-Luce likelihood function with additional
reified entities.

@@

%% BibTeX support
\bibliographystyle{apalike2}
\begin{thebibliography}{}

\bibitem[Bakhrankova, 2011]{bakhrankova2011}
Bakhrankova, K. (2011).
\newblock Points scoring systems in {F}ormula 1: A game theoretic perspective.
\newblock In D. Klatte (Ed.), {\em Operations Research Proceedings}  (pp.\
  211--216).: Springer-Verlag.

\bibitem[Bradley, 1952]{bradley1952}
Bradley, R.~A. (1952).
\newblock Incomplete block rank analysis: On the appropriateness of the model
  for a method of paired comparisons.
\newblock {\em Biometrics}, 10(3), 375--390.

\bibitem[Codling, 2017]{codling2017}
Codling, S. (2017).
\newblock {\em Speed read {F1}}.
\newblock Motorbooks.
\newblock ISBN 978-0-7603-5562-6.

\bibitem[Emerson, 2007]{emerson2007}
Emerson, P. (2007).
\newblock {\em Designing an All-Inclusive Democracy---Consensual Voting
  Procedures for use in Parliaments, Councils and Committees}.
\newblock Springer-Verlag.
\newblock {I}{S}{B}{N} 978-3-540-33163-6.

\bibitem[Gay-Rees, 2019]{gayrees2019}
Gay-Rees, J. (2019).
\newblock {F}ormula 1: drive to survive.
\newblock Netflix documentary.

\bibitem[Hankin, 2010]{hankin2010}
Hankin, R. K.~S. (2010).
\newblock A generalization of the {D}irichlet distribution.
\newblock {\em Journal of Statistical Software}, 33(11), 1--18.

\bibitem[Hankin, 2017]{hankin2017}
Hankin, R. K.~S. (2017).
\newblock Partial rank data with the {hyper2} package: likelihood functions for
  generalized {B}radley-{T}erry models.
\newblock {\em The R Journal}, 9(2), 429--439.

\bibitem[Jenkins, 2010]{jenkins2010}
Jenkins, M. (2010).
\newblock Technological discontinuities and competitive advantage: a historical
  perspective on {F}ormula 1 motor racing 1950-2006.
\newblock {\em Journal of Management Studies}, 47(5), 884--910.

\bibitem[Luce, 1959]{luce1959}
Luce, R. (1959).
\newblock {\em Individual choice behaviour: A theoretical analysis}.
\newblock New York: Wiley.

\bibitem[Mastromarco \& Runkel, 2009]{mastromarco2009}
Mastromarco, C. \& Runkel, M. (2009).
\newblock Rule changes and competitive balance in {F}ormula {O}ne motor racing.
\newblock {\em Applied Economics}, 41(23), 3003--3014.

\bibitem[Mollica \& Tardella, 2014]{mollica2014}
Mollica, C. \& Tardella, L. (2014).
\newblock Epitope profiling via mixture modeling of ranked data.
\newblock {\em Statistics in Medicine}, 33, 3738--3758.

\bibitem[Plackett, 1975]{plackett1975}
Plackett, R.~L. (1975).
\newblock The analysis of permutations.
\newblock {\em Applied Statistics}, 24, 193--202.

\bibitem[Wood, 2020]{wood2020}
Wood, E. (2020).
\newblock {F}ormula {S}cout (published online).

\bibitem[Zipf, 1949]{zipf1949}
Zipf, G. (1949).
\newblock {\em Human Behavior and the Principle of Least Effort}.
\newblock Addison-Wesley.

\end{thebibliography}

>>>>

\documentclass{elsarticle}

\usepackage{lineno,hyperref}
\usepackage{amsmath}
\usepackage{mathptmx}
\usepackage{graphics}
\usepackage{setspace}\doublespacing
\modulolinenumbers[5]

\journal{Journal of Economic Behavior and Organization}

\bibliographystyle{model2-names.bst}\biboptions{authoryear}

@@

\begin{frontmatter}
  \title{An objective assessment of point scoring systems in Formula 1 motor racing}
  \author{Robin K. S. Hankin\fnref{myfootnote}}
\address{AUT University, 55 Wellesley Street, Auckland 1010, New Zealand}

\begin{abstract}
The points scoring system of Formula 1 motor racing is a long-standing
contention among fans and competitors.  The inaugural points system
was $(8,6,4,3,2)$, that is, 8 points to the winner, 6 to second place,
and so on; in 2020 the points system was changed to
$(25,18,15,12,10,8,6,4,2,1)$.  However, it is difficult to assess
statistical nulls using accumulated points.  Here, I use Plackett-Luce
likelihood to identify a ranking for the competitors from the
perspective of tournament theory, this method being amenable to
statistical testing.  I go on to assess a number of reasonable points
systems objectively: in one well-defined statistical sense a Borda
points system is optimal.
\end{abstract}
\begin{keyword}
Reified Bradley-Terry\sep Likelihood\sep Formula 1 motor racing\sep Points systems
\end{keyword}

\end{frontmatter}
\linenumbers

@@

\section{Introduction}\label{introduction}
Formula 1 motor racing is an important and prestigious motor sport
\citep{codling2017,jenkins2010}.  Season ranking is based on a points
allocation system wherein competitors are awarded points based on race
finishing order; points accumulate additively.  The overall
competition winner is the competitor who accumulates the most points
after the final race.  The intent of the points system is to
incentivize competitors, stimulate innovation, and to create an
exciting sporting spectacle: as such, its study is a practical
application of tournament theory~\citep{lazear1981}.

@@

(observe how this likelihood function, while informative about
$p_4+p_5$, is uninformative about $p_4\left|p_4+p_5\right.$). We now
use a technique due to \cite{hankin2010,hankin2020} and introduce
fictional (reified) entities whose nonzero Bradley-Terry strength
helps certain competitors or sets of competitors under certain
conditions. The canonical example would be the home-ground advantage
in association football.  If players (teams) $1,2$ with strengths
$p_1,p_2$ compete, and if our observation were $a$ home wins and $b$
away wins for team $1$, and $c$ home wins and $d$ away wins for
team~$2$, then a suitable likelihood function would be

@@

\noindent where $p_H$ is a quantification of the beneficial home
ground effect.  Similar techniques have been used to account for the
first-move advantage in chess, and effective coordination between
members of doubles tennis teams; we may use a similar device to
account for (e.g.) wet conditions in Formula 1.  Here I analyse
seasons 2016-2019 using the \texttt{hyper2} package \citep{hankin2017}
which implements the Plackett-Luce likelihood function with additional
reified entities~\cite{hankin2020}.

@@

%% BibTeX support
\bibliography{formula1}
